---
title: "paralogSL: De Novo Paralog Synthetic Lethality Discovery"
author: "Tao Zhu"
date: "2026-05-25"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{paralogSL: De Novo Paralog Synthetic Lethality Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(paralogSL)
library(ggplot2)
```

## Overview

`paralogSL` provides a lightweight R framework for de novo synthetic lethality
(SL) discovery through paralog compensation analysis. It computes Delta
Dependency (DD) scores from DepMap CRISPR dependency data and identifies
paralog-SL candidates using HGNC gene family annotations.

Key features:
- **DD computation**: One-line subtraction; interpretable by construction
- **Cross-cancer**: Built-in driver gene sets for 5 cancer types
- **Benchmark**: Pre-loaded contextual comparison against 8 published methods
  (not a head-to-head benchmark — see `?benchmark_methods`)
- **Visualization**: Publication-quality plots with Chinese-style palette

## Quick Start

```{r}
# Load built-in data
data("gyn_drivers")
data("known_sl_pairs")
data("benchmark_methods")

# View driver genes
str(gyn_drivers)

# View gold standard
head(known_sl_pairs)
```

## Loading DepMap Data

```{r eval=FALSE}
# Download from https://depmap.org/portal/download/
dep <- load_dependency("CRISPRGeneEffect.csv")
expr <- load_expression("OmicsExpressionProteinCodingGenesTPMLogp1.csv")
models <- load_models("Model.csv")
mutations <- data.table::fread("OmicsSomaticMutations.csv")
```

## Computing Delta Dependency

```{r eval=FALSE}
# Filter to ovarian cancer cell lines
ov_cells <- filter_cancer_cell_lines(models,
    c("Ovarian Serous Cystadenocarcinoma",
      "Ovarian Clear Cell Adenocarcinoma",
      "Ovarian Endometrioid Adenocarcinoma"))

# Load HGNC paralog pairs
paralogs <- data.table::fread("ensembl_paralogs.csv")

# Run analysis
results <- run_paralog_analysis(
    dependency = dep,
    expression = expr,
    mutations_df = mutations,
    paralogs_df = paralogs,
    driver_genes = gyn_drivers$Ovarian,
    cell_lines = ov_cells
)

head(results, 10)
```

## Computing AUROC

```{r eval=FALSE}
auroc <- compute_auroc(results, known_sl_pairs)
cat("DD AUROC:", round(auroc, 3), "\n")
```

## Benchmark Comparison

```{r}
bench <- get_benchmark_table()
print(bench)

plot_benchmark(bench)
```

## Individual DD Computation

```{r eval=FALSE}
# Build mutation matrix for ARID1A
mut_mat <- build_mutation_matrix(mutations, ov_cells, "ARID1A")
arid1a_mut <- names(which(mut_mat[, "ARID1A"] == 1))
arid1a_wt  <- names(which(mut_mat[, "ARID1A"] == 0))

# Compute DD for ARID1A → ARID1B
dd_result <- compute_dd(dep, "ARID1A", "ARID1B", arid1a_mut, arid1a_wt)
cat("DD:", dd_result$DD, "p:", dd_result$p_value, "\n")
```

## CPTAC Protein Correlation Plot

```{r eval=FALSE}
prot <- read.table("FD_GLBL_MI_FFPEbridge_Abund_20201002.tsv",
                   sep = "\t", header = TRUE, row.names = 1)

plot_protein_correlation(prot, "ARID1A", "ARID1B")
```

## MSI Stratification

```{r eval=FALSE}
# Classify cell lines by MSI/dMMR status
msi_status <- classify_msi_status(mutations, ov_cells)
msi_h_cells <- names(msi_status[msi_status == "MSI_H"])
mss_cells   <- names(msi_status[msi_status == "MSS"])

# Run DD analysis separately for MSI-H vs MSS
msi_h_results <- run_paralog_analysis(dep, expr, mutations,
    paralogs, gyn_drivers$Colorectal, msi_h_cells)
mss_results   <- run_paralog_analysis(dep, expr, mutations,
    paralogs, gyn_drivers$Colorectal, mss_cells)

# Compare AUROC
msi_h_auc <- compute_auroc(msi_h_results, known_sl_pairs)
mss_auc   <- compute_auroc(mss_results, known_sl_pairs)
cat("MSI-H AUROC:", msi_h_auc, "\nMSS AUROC:", mss_auc, "\n")
```

## Therapeutic Window Analysis

```{r eval=FALSE}
# Compute therapeutic index for ARID1A → ARID1B
tw <- compute_therapeutic_window(
    dependency = dep,
    driver_gene = "ARID1A",
    paralog_gene = "ARID1B",
    mut_lines = arid1a_mut,
    wt_lines = arid1a_wt,
    all_cell_lines = ov_cells
)
cat("TI:", tw$TI, "\n")
cat("Classification:", tw$classification, "\n")
cat("Selectivity:", tw$selectivity, "\n")
```

## Mutation Type Stratification

```{r eval=FALSE}
# Stratify BRCA1 mutations by type
mt <- stratify_by_mutation_type(mutations, ov_cells, "BRCA1")
cat("Truncating mutants:", length(mt$trunc_lines), "\n")
cat("Missense mutants:", length(mt$miss_lines), "\n")

# Compute DD separately for each mutation type
dd_trunc <- compute_dd(dep, "BRCA1", "BRCA2", mt$trunc_lines, arid1a_wt)
dd_miss  <- compute_dd(dep, "BRCA1", "BRCA2", mt$miss_lines, arid1a_wt)
cat("DD (truncating):", dd_trunc$DD, "\nDD (missense):", dd_miss$DD, "\n")
```

## Reproducibility

This vignette reproduces the core findings from the manuscript:

> *Delta Dependency Prioritizes Paralog-Based Synthetic Lethality
> Candidates Across Solid Tumor Types* (Mo & Zhu, Genome Biology, 2026)

Current headline evaluation (paralog-sl-predictor v1.1.0 artifacts):

- DD AUROC = 0.676 on the full lineage-level frame (110 driver--paralog--lineage
  entries, 8 positives); 7 of 8 evaluable lineages reach AUROC >= 0.7
- DD AUROC = 1.000 on the Tier A + Tier B external benchmark (2 of 5 pairs
  lineage-evaluable; evidence-tiered gold standard, `?known_sl_pairs`)
- Pair-level aggregate: DD alone 0.566; interpretable composite score 0.831
  (AUPRC 0.356, 4.3x baseline prevalence)
- DD + 30% k-mer identity filter -> AUROC = 1.000 on a 3-pair high-identity
  subset (2 positives; anecdotal)
- Published CV3 references (e.g. SLMGAE 0.790, Feng et al. 2024) are
  contextual values from a general SL universe, NOT a head-to-head benchmark
- CPTAC protein co-variation supports compensation at the protein level

`compute_auroc()` defaults to the Tier A + Tier B primary benchmark
(`tiers = c("A", "B")`); pass `tiers = NULL` to reproduce the pre-v1.1.0
all-pairs behavior.

See the manuscript and Python pipeline at
\url{https://github.com/tjogzt/paralog-sl-predictor} for full details.
